# Cross Recurrence Quantification Analysis

Some text ...

An inline equation can use LaTeX mathematics, e.g., $f(t) = t^2 e^-2t$.

It is also possible to have display equations that take up a whole line:
$$
  z_n = z_{n + 1}^2 + c.
$$

```{r}
#| echo: false

# Note that this code block is hidden to the user, since echo is set to false
# in the special comment above which is a code chunk option.

numbers <- 1:100
```

Consider the numbers `{r} numbers`.

You can easily sum these (stored in the variable `numbers`) using R:

```{r}
summed_numbers <- sum(numbers)
```

If you are curious, the sum is `{r} summed_numbers`, something the young
Carl Friedrich Gauss, allegedly,
[solved](https://www.americanscientist.org/article/gausss-day-of-reckonin) in
class as a young schoolboy.

Plotting the sine of these numbers mapped to the interval $[0, 2\pi$ is
undoubtedly a much more daunting task to perform manually, but --- again --- we
can ask R to do it for us:

```{r}
library(ggplot2)
ggplot(data.frame(x = numbers * 2 * pi / 100),
       aes(x = x)) +
  geom_function(fun = sin) +
  theme_classic()
```

